---
output:
  html_document: default
  pdf_document: default
---
## Assignment 1
```{r, echo = F}
print("Sebastian Hyland (39878921)")
```


### Dataset overview
Origin: part of the data published by [Blackmore et al. (2017)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5544260/)

Study: The effect of upper-respiratory infection on transcriptomic changes in the CNS

File: CSV format:

| Column                                | Description                                                  |
| ------------------------------------- | ------------------------------------------------------------ |
| gene                                  | The name of the gene that was measured                       |
| sample                                | The name of the sample the gene expression was measured in   |
| expression                            | The value of the gene expression                             |
| organism                              | The organism/species - here all data stem from mice          |
| age                                   | The age of the mouse (all mice were 8 weeks here)            |
| sex                                   | The sex of the mouse                                         |
| infection                             | The infection state of the mouse, i.e. infected with Influenza A or not infected. |
| strain                                | The Influenza A strain; C57BL/6 in all cases.                |
| time                                  | The duration of the infection (in days).                     |
| tissue                                | The tissue that was used for the gene expression experiment, i.e. cerebellum or spinal cord. |
| mouse                                 | The mouse unique identifier.                                 |
| ENTREZID                              | The gene ID for the ENTREZ database                          |
| product                               | The gene product                                             |
| ensembl_gene_id                       | The ID of the gene from the ENSEMBL database                 |
| external_synonym                      | A name synonym for the gene                                  |
| chromosome_name                       | The chromosome name of the gene                              |
| gene_biotype                          | The gene biotype                                             |
| phenotype_description                 | The phenotype description of the gene                        |
| hsapiens_homolog_associated_gene_name | The human homologous gene                                    |


\pagebreak


### Processing and Analysis
First, load the data:
```{r}
data <- read.csv(file="rna_seq.csv")
```

**Question 1)** Count the number of unique genes with expression levels greater than the median expression of the dataset. Ensure you exclude any missing values (NA) from the expression data before performing your calculations.

First, let us calculate the median expression level (removing missing values) and subset the dataframe to rows with above-median expression levels:
```{r}
median_expression <- median(data[["expression"]], na.rm = TRUE)
above_median_expression_subset <- subset(data, expression > median_expression)
```
Now, we can count the number of rows with a unique "gene" identifier:
```{r}
unique_genes <- length(unique(above_median_expression_subset[["gene"]]))
print(unique_genes)
```


\pagebreak



**Question 2)** Write code to determine the number of unique mice (based on the mouse column) that were infected (infection == "InfluenzaA") and are of the female sex (sex == "Female").

Again, we can subset to the dataset to only rows that match infection == "InfluenzaA" AND sex == "Female":
```{r}
female_infected_subset <- subset(
  data,
  infection == "InfluenzaA" & sex == "Female"
)
```
and then count the number of rows with a unique "mouse" identifier
```{r}
unique_mice <- length(unique(female_infected_subset[["mouse"]]))
print(unique_mice)
```


\pagebreak


**Question 3)** Find the number of common genes that are expressed in both male and female mice. This means identifying genes that have expression values greater than zero for both sexes.

We can first subset the data to only rows with expression values greater than zero, then subset again to split by sex. For each sex-based subset, we will take only the "gene" column:
```{r}
expressed_subset <- subset(data, expression > 0)
male_unique_subset <- unique(
  subset(expressed_subset, sex == "Male")[["gene"]]
)
female_unique_subset <- unique(
  subset(expressed_subset, sex == "Female")[["gene"]]
)
```
Now that we have a vector of genes with an expression level greater than zero for each sex, we can intersect the two sets, and find its length, giving us the number of common genes expressed in both male and female mice:
```{r}
common_genes <- length(
  intersect(male_unique_subset, female_unique_subset)
)
print(common_genes)
```


\pagebreak


**Question 4)** For each mouse, identify the gene with the lowest expression level on chromosome 1, ensuring that the expression value is greater than zero. The final result should be a dataframe with two columns: mouse and lowest_gene. You might use `rbind()` to add the results iteratively.

First, we will subset the data to rows with an expression level greater than zero, and a gene that is part of chromosome 1:
```{r}
chromosome_one_subset <- subset(data, chromosome_name == 1 & expression > 0)
```
Next, we can split this subset into a list, where each element is a data frame containing information concerning a unique mouse. Then, we can find the element with the lowest expression, and re-combine all list elements into a singular dataframe:
```{r}
list_per_mouse <- split(chromosome_one_subset, chromosome_one_subset[["mouse"]])
lowest_expression_genes <- lapply(list_per_mouse, function(mouse_df) {
  mouse_df[which.min(mouse_df[["expression"]]), c("mouse", "gene")]
})
recombined_df <- do.call(rbind, lowest_expression_genes)
```
Finally, we can rename columns to their expected names:
```{r}
colnames(recombined_df) <- c("mouse", "lowest_gene")
print(recombined_df, row.names = FALSE)
```


\pagebreak


**Question 5)** Write a function that finds the top 3 genes with the highest expression levels for each chromosome in the dataset. You can utilize `list()` data structure and `order()` function.

We can define a function that splits the dataframe into a list of unique chromosomes, orders each list element's dataframe by expression level in decreasing (higher-first) order, takes the first 3 rows, and rejoins the list into a singular dataframe:
```{r}
get_top_3_chromosomes <- function(input_df) {
  list_per_chromosome <- split(input_df, input_df[["chromosome_name"]])
  list_per_chromosome_top3 <- lapply(
    list_per_chromosome, function(chromosome_df) {
      chromosome_df[
        order(
          chromosome_df[["expression"]],
          decreasing = TRUE
        )[1:3],
        # Keep these columns
        c("chromosome_name", "gene", "expression")
      ]
    }
  )
  recombined_chromosome_df <- do.call(rbind, list_per_chromosome_top3)
  recombined_chromosome_df
}
```
Now, we can call the function on our input dataframe, and print the result
```{r}
result <- get_top_3_chromosomes(data)
print(result, row.names = FALSE)
```


\pagebreak


**Question 6)**  To examine the variations in gene expression between males and females, create a Violin plot illustrating the distribution of gene expression levels, separating the data by both **sex** and **infection** status. The y-axis should be on a logarithmic scale (log10). Provide meaningful labels for the axes and ensure the plot is colored by **infection** status.

First, let us import the `ggplot2` and `nord` libraries, and interpret `sex` and `infection` as categorical factors
```{r}
library(ggplot2); library(nord)
sex_factor <- as.factor(data$sex)
infection_factor <- as.factor(data$infection)
```
Now, let us plot the data, displaying sex on the x-axis, gene expression on the y-axis (on a logarithmic scale), and infection as colour:
```{r, warning=FALSE, fig.height=4}
time_factor <- as.factor(data$time)
infection_factor <- as.factor(data$infection)
sex_factor <- as.factor(data$sex)
ggplot(
  data,
  aes(x = time_factor, y = expression, fill = sex_factor, theme = "dark")
) +
  theme_bw() +
  labs(
    title = "Gene expression levels based on sex and infection time",
    y = "Expression (" ~ log[10] ~ "scale)",
    x = "Infection time (days)",
    fill = "Sex"
  ) +
  geom_violin(scale = "count") +
  scale_y_continuous(transform = "log10") +
  scale_fill_nord("algoma_forest")
```
